
# 6. Analysis of longitudinal data

In this chapter I perform analysis of longitudinal data based on datasets RATS and BPRS. The analysis ought to follow those in the book Multivariate Analysis for the Behavioral Sciences (MABS) for chapters 8 and 9. Before the analysis, some data wrangling is involved.

## Data Wrangling

Before analysis, the data is manipulated from wide form to long form. The analysis is performed in the R-file "meet_and_repeat.R" that can be found in the github repository. Also, the datasets are described there.

## Analysis 1: Chapter 8 of MABS using the RATS data

As suggested earlier, the analysis follows that of chapters 8 and 9 in the book MABS. The first part deals with dataset RATS and Chapter 8 of MABS.

### Dataset RATS explained

RATS dataset is from Crowder and Hand (1990), who studied the effects of of different diets on the weight of three groups of rats.
As shown below, using the dataset in the long form, the variables are ID (number associated with each rat), Group (associated with one of the three diets), WD (the initial time of observation), Weight (the weight of the rat measured at different points of time) and Time (the time of observation).



```{r}
rm(list=ls()) 
library(dplyr)
library(stringr)
library(ggplot2)
library(GGally)
library(tidyr)
library(lme4)
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", stringsAsFactors = F, na.strings = "..")
RATSL<-read.table("RATSL.txt")
head(RATSL)
str(RATSL)
summary(RATSL)
glimpse(RATSL)
```


### Graphical displays of longitudinal data and comparison of groups

This section focuses on displaying the data in a way that something can be learned.
First, lets plot the data by groups.


```{r}
# Plot the RATSL data by groups
ggplot(RATSL, aes(x = Time, y = Weight,col=Group, linetype = as.factor(ID)))+geom_line()+scale_linetype_manual(values = rep(1:10, times=4))+facet_grid(. ~ Group, labeller = label_both) +theme(legend.position = "none")
```

As suggested by the plots, in all groups and for all individuals, the weight seems to increase in time. What stands out in the plots is that there are big differences in the weights of the rats in different groups such that those in group 1 are the lightest and those in the group are the heaviest with the exception of one sturdy rat in group two that is a clear outlier. 

Overall, the plot support the tracking phenomenom mentioned in MABS that the heavier rats is in the beginning are often the heaviest in the end as well.

Next, let's study the same thing after standardizing the data. In the standardized dataset, the weights are scaled according to (weight - mean(weight)) / se(weight).

```{r}
# Scale RATSL, new data RATSL1
RATSL1 <- RATSL %>%
group_by(Time) %>%
mutate(scaled_weight=(Weight-mean(Weight))/sd(Weight)) %>%
ungroup()
glimpse(RATSL1)
# Plot scaled data by groups
ggplot(RATSL1, aes(x = Time, y = scaled_weight, group=ID, col=Group, linetype=as.factor(ID)))+geom_line()+scale_linetype_manual(values = rep(1:10, times=4))+facet_grid(. ~ Group, labeller = label_both) +theme(legend.position = "none")
```

From the scaled data it appears that the growth rates seem to be the same among groups.

Next, means and standard error of groups are studied.

```{r}
n <- RATSL1$Time %>% unique() %>% length()
RATSL1$Group<-factor(RATSL1$Group)
RATSL2 <- RATSL1 %>%
group_by(Group, Time) %>%
summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
ungroup()
#Plot profiles
ggplot(RATSL2, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
geom_line() +
geom_point(size=2) +
geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.5) +
theme(legend.position = c(0.9,0.6)) +
scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

The plot above suggests that the second group has the highest variation, which is probably driven by the sturdy outlier rat. The second highest variation variation was in the third group and the smallest in the first group. The mean weights grew in all groups and the highest mean is in the third group, second highest in the second group and the smallest in the first group.

### Summary measure analysis of longitudinal data

The summary measure analysis of the rats is performed with boxplots. To make the comparison more meaningful, the sturdy outlier rat from group 2 is filtered. 

```{r}
RATSL3 <- RATSL %>%
filter(Time > 1) %>%
group_by(Group, ID) %>%
summarise(mean=mean(Weight) ) %>%
ungroup()
#look at the data
glimpse(RATSL3)
#structure of the data
str(RATSL3)
#remove the fat ourlier from group 2
RATSL4<-filter(RATSL3, mean<550)
ggplot(RATSL4, aes(x = as.factor(Group), y = mean)) +
geom_boxplot() +
stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
scale_y_continuous(name = "mean(Weight)")
```

There still seems to be one outliers in groups 1 and 3. This may explain why the variances in groups 1 and 3 are greater than in group 2. As noted earlier, the groups vary a lot in terms of means.

Last step here is to study if the differences between these three groups are statistically significant, which is done using analysis of variance (ANOVA) (for 2 groups one could use two sample t-tests). Assumptions in ANOVA are that the observations are independent from each other, the data in each group is normally distributed and they have a common variance.

In ANOVA, h0 is that the means of the group don't differ from each other, which is tested both for the baseline scenario at T=1 and for T>1.

```{r}
RATS1 <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt")
RATSL5 <- RATSL3 %>% mutate(Baseline = RATS1$WD1)
fit <- lm(mean ~ Baseline + Group, data = RATSL5)
anova(fit)
```

ANOVA suggests that the groups differ from each other only at the Baseline, which means that in the beginning, the groups are statistically different (h0 rejected). Then again, there are no differences during the study at 5% significance level as h0 can't be rejected. 

## Analysis 2: Chapter 9 of MABS using the BPRS data
The second part of analysis deals with dataset BPRS and Chapter 9 of MABS.

### Dataset BPRS explained

BPRS dataset is from a study by Davis (2002), who measured a brief psychiatric rating score (BPRS) from 40 males, who were assigned to two different groups, which got different treatments. The BPRS was measured once a week for eight weeks and the BPRS score was used for studying patients who may have schizophrenia. 

Below is a numerical look into the dataset, from which we see the information as explained: treatment tells the treatment group (1 or 2) of each individual, subject is the number associated with each of the participants, bprs is the measured bprs score for each individual for each week and week is associated with the time of the measured bprs score.


```{r}
BPRSL<-read.table("BPRSL.txt")
head(BPRSL)
str(BPRSL)
summary(BPRSL)
BPRSL$treatment <- factor(BPRSL$treatment)
```

Let's begin with plotting the data by the groups.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = as.factor(subject))) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

Not much can be said based on the plots for either of the groups. It seems that the BPRS seems to decrease in time, especially for group 1 on the left. Having so many individuals makes the interpretation based on the plots quite difficult.

### Basic linear model 

Without taking into account the structure of the data, which includes many observations for each individual, regular linear regression is performed.

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

Here, treatment is not statistically significant while week is, which is in line with the previous observation of possible negative trend.

### Linear Mixed Effects models

As the regular linear model is not very informative in this case because of the data, intercept is allowed to vary for each individual in the random intercept model below. This is something very nice called linear mixed effects model. Later on there are more variations to this.

```{r}
bprsl_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(bprsl_ref)
```

From the summary one can see the variances of subjects and the correlation of fixed effects. Compared to the regular linear model, the coefficients are the same, but the standard errors for "week" and "treatment" are smaller.

Next, both random intercept and random slope are used in the linear model. This allows for the fit of the model for each individual in the data to differ in intercept and in slope. This is used to allow the model to adjust for the individual differences in the subjects's bprs values and for the different developments during the period.

```{r}
#Linear mixed effects model with random slope and random intercept
BPRS_ref1<- lmer(bprs ~ week + treatment +(week | subject), data=BPRSL, REML=FALSE)
summary(BPRS_ref1)
```

Lastly, the LME model is performed with random slope and random intercept and time*group interaction.

```{r}
#Linear mixed effects model with random slope and random intercept and time*group interaction
BPRS_ref2<- lmer(bprs ~ week*treatment+(week | subject), data=BPRSL, REML=FALSE)
summary(BPRS_ref2)
```

ANOVA for random intercept random slope model vs the random slope and random intercept and time*group interaction model. ANOVA suggests that the lower the p-value, the better the fit against the comparison model. Now, depending on where we set the significance level, let's say it's 10%, we can conclude that the model with the interaction term (BPRS_ref2) has better fit than the model without it (BPRS_ref1).

```{r}
anova(BPRS_ref2, BPRS_ref1)
```

Following the DataCamp exercise, the last thing in this chapter is to plot the actual data and the fitted data for both groups (treatment = 1 (LHS) vs treatment = 2 (RHS)). Like earlier, each individual is represented by it's own line (and color).

```{r}
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
BPRSL <- mutate(BPRSL, subject2 = ifelse(BPRSL$treatment==2, BPRSL$subject+20, BPRSL$subject)) 


ggplot(BPRSL, aes(x = week, y = bprs, group = subject2,col= subject)) + 
  geom_line(aes(linetype = treatment)) +  
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Time", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "bottom")

ggplot(BPRSL, aes(x = week, y = Fitted, group = subject2, col= subject)) +
  geom_line(aes(linetype = treatment)) +  
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Time", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "bottom")

```
